# Vrije opdracht

## Keuze onderwerp
Ik heb er voor gekozen om mijzelf te verdiepen in het analyseren van single cell RNA-sequencing data.


## Theoretische achtergrond
### Wat is single cell RNA-sequncing
Single cell RNA-sequencing (scRNA-seq) is een methode om verschillende celtypes in een heterogeen monster/biopt te identificeren. Voor deze techniek is zowel de praktische uitvoering als de juiste analyse van de data essentieel
om verschillende celtypes te kunnen identificeren.

In het praktische deel word een biopt of ander type monster (b.v. hersenweefsel of bloed). Het biopt word gesuspendeerd zodat alle cellen los van elkaar in de oplossing zitten. Vervolgens is het de bedoeling dat al het RNA in één cel een unieke barcode krijgt, en dat elk RNA fragment in één cel een Unique Molecular Identifier (UMI) krijgt. De barcode kan worden gebruikt om al het RNA uit één cel te kunnen identificeren. De UMI is belangrijk om te bepalen hoeveel (na amplificatie d.m.v. PCR) unieke fragmenten er zijn en hoe vaak deze geamplificeerd zijn. Zonder UMI kan je bijvoorbeeld niet achterhalen of 1 frament 100 keer is geamplificeerd tijdens PCR, of dat er 100 van deze fragmenten al aanwezig waren in de cel.

Deze barcode en UMI worden aan een cel gekoppeld via beads die gecoat zijn met RNA sequenctie die de barcode en UMI bevatten. Hiervoor is speciaal microfluidics aperatuur ontwikkeld die er voor zorgen dat één cel en één bead in een klein olie druppeltje terecht komen. In de druppel word de cel gelyseerd, binden de RNA fragmenten van de cel aan de RNA fragmenten op de bead en vind er reverse transcriptie plaats (cDNA synthese). Vervolgens worden de fragmenten geamplificeerd d.m.v. PCR en tot slot worden de fragmenten gesequenced met b.v. een illumina sequencer. Voor het sequencen krijgen de cDNA fragmenten sequencing adapters die nodig zijn om te binden bij illumina sequencing. Dit is het einde van de praktische uitvoering van scRNA-seq. De rest van de analyse word digitaal uitgevoerd met bijvoorbeeld R.

### Anylseren van single cell RNA-sequencing data in R
De output van het sequencen zijn FASTQ bestanden

## Planning
- learn how to use the seurat package using a tutorial (20 uur)
- perform the analysis on real sequencing data from an experiment (12 uur)

## Uitvoering op dataset
The dataset I have chosen to perform the analysis on is one from the Human Immune Health Atlas, found on CELLxGENE. They have  many datasets on specific cells and I have chosen one about CD4 T cells because of my interest in immunology.


```{r 03 libraries, include=F}
library(Seurat)
library(patchwork)
options(Seurat.object.assay.version = "v3")
```

### Step 1. Create a Seurat object
First a seurat object has to be creating using the data. This is an introduced datatype which stores results during the analysis. Because my data is of the .h5ad type I first need to convert it so its compatable with seurat. I'm using the SeuratDisk package for this.

```{r Unpack data, eval=F}
untar("data/scrna_data/pbmc_unsorted_10k_filtered_feature_bc_matrix.tar.gz", 
      exdir = "data/scrna_data/")
```

```{r create seurat object}
counts <- Read10X(data.dir = "data/scrna_data/filtered_feature_bc_matrix/")

pbmc <- CreateSeuratObject(counts, project="pbmc10k")
```


### Step 2. Quality control
The next step is to do a quality control by applying filters to the dataset. These filters may include removing Cells with too few or too many genes detected and removing cells with high mitochondrial transcript percentage which may indicate stress. This last filter needs to be calculated manually. 

First we check if the distribution of the different features is as expected. nFeature_RNA is the number of unique detected genes and nCount_RNA is the total number of transcripts. the percent.mt is the percentage of transcripts that come from mitochondrial genes per cell. A low percentage (0-20) indicates the cells are healthy.

The graph shows that both nFeature and nCount are high. If the transcript count is high that should mean that a lot of genes are detected aswell, meaning the dsitribution looks normal. Percent.mt is below 5 percent which indicates the cells did not experience stress.

```{r Visualise distribution, warning=F}
# Calculate mitochondrial transcirpt percentage
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT[-\\.]")

# Visualise transcript distribution
VlnPlot(pbmc,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol=3,
        pt.size=0,
        cols = "#41b6c4")
```

To decide where to cutoff the data the Ncount_RNA is plotted against the other two features. Most cells seem to have a mitochondrial transcript expresion below 10%. Because I dont want to cut out too many cells I've decided to choose 15% as the cutoff value. The correlation between nFeature and nCount is quite high (0.97) so not a lot of trimming needs to be done. I've decided for a cutoff value of 30k.

```{r feature scatter, fig.width=10, fig.height=6, out.width="100%"}
# Plot distribution
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt", cols = "#2c7fb8")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", cols = "#a1dab4")
plot1 + plot2
```

```{r data cutoff}
# Cutoff data
pbmc <- subset(pbmc, subset =
                 nFeature_RNA > 500 &
                 nFeature_RNA < 30000 &
                 percent.mt <15)
```

### Step 3 Normalization
Because the number of RNA transcripts can differ per cell, the data has to be normalized so the RNA read count actually represents the expression of a gene compared to other cells.

```{r normalize data}
pbmc <- NormalizeData(pbmc)
```
## Step 4. Feature selection for following heterogeneity analysis
Different cell types can be identified based on their distinct expression signature. To do this you have to look at the genes which are most variably expressed across cells and filter out genes with low expression levels or genes which are expressed consistently across most celltypes. These highly variable genes are called features.

There is no rule of thumb to choose the feature count, but usually a featurecount between 2000 and 5000 will obtain good results. Because this is a well defined dataset, I'm choosing 200 features whihc is seen as the standard.

```{r Select features, results='hide'}
pbmc <- FindVariableFeatures(pbmc, nfeatures = 2000)
```

```{r plot top features, warning=FALSE, out.width="100%", fig.width=10, fig.height=6}
top_features <- head(VariableFeatures(pbmc), 10)
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top_features, repel = TRUE)
plot1 + plot2
```

## Step 5. Data scaling
Because some genes are more highly expressed they will seem to contribute more to the analysis even though all genes should have the same contribution. To solve this the data gets scaled. Additionally unwanted sources of variation can be removed. It is advised to only do this after obtaining the first results and checking for unwated variation.

## Bronnen
https://www.10xgenomics.com/blog/how-does-single-cell-rna-seq-work
https://www.10xgenomics.com/blog/how-single-cell-sequencing-data-analysis-works
https://satijalab.org/seurat/articles/pbmc3k_tutorial
https://cellxgene.cziscience.com/datasets
